Mergeable Types

### 2017-02-28 paper/examples.tex
edits

\noindent The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The implementation
		specified in Fig.~\ref{fig:mlist-sig} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the second element from $v_2$, if it were to
		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
		\C{op\_transform} (shown in the Appendix) computes such
		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
		edit sequence $s_2$. A notable transformation rule is for a substitution
		w.r.t another substitution at the same position. Since the substituting
		items could be different, the function relies on the item merge function
		(\C{A.merge}) to merge them into a single item. For instance, in
		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

>>>>

\noindent The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml.  The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The signature
		specified in Fig.~\ref{fig:mlist} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the second element from $v_2$, if it were to
		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
		\C{op\_transform} (shown in the Appendix) computes such
		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
		edit sequence $s_2$. A notable transformation rule is for a substitution
		w.r.t another substitution at the same position. Since the substituting
		items could be different, the function relies on the item merge function
		(\C{A.merge}) to merge them into a single item. For instance, in
		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

### 2017-02-28 paper/related.tex
edits

There have been numerous proposals over the years for realizing distributed
functionality in a functional programming context. Poly/ML~\cite{Mat97} and
Facile~\cite{TLK+93} extend Standard ML, while Acute~\cite{SLW+05} and
HashCaml~\cite{BSS+06} extend OCaml, with distributed programming abstractions
and support, with particular focus on type-safe marshalling and mobility.
Kali-Scheme~\cite{CJK95} defines a distributed extension for Scheme-48, and
Cloud Haskell~\cite{EBPJ11} describes a domain-specific language that enables
development of distributed Haskell programs. Obliq~\cite{Car95} explores issues
related to mobility, lexical scoping, and network references in an
object-oriented framework that enables safe (i.e., lexical) transmission of
closures. Orleans~\cite{Orleans} provides persistent actor-based programming
model for elastic cloud application, but the focus is on strong consistency,
mirroring a single-threaded execution.

However, unlike \name, none of these systems consider issues of object
replication or consistency as central to their programming models. The
realities of modern-day scalable distributed systems, especially those that are
constructed from geographically distributed participants, dictate that we must
necessarily grapple with consistency issues to achieve any kind of reasonable
performance; it is noteworthy that replication is inherent in many widely-used
cloud-based distributed system implementations today (e.g.,
Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide only weak
consistency guarantees.  Rather than exposing a low-level operational treatment
of replication, foisting the burden of reasoning about replicated state onto
the programmer, our primary contribution is the development of a principled
high-level approach to framing these issues that abstracts these kinds of
system-level details.  Our goal is to leverage functional programming
principles to minimize disruption to the way programmers structure and reason
about their applications, without compromising efficiency or performance.
Indeed, the semantics of mergeable types and versioning has utility in any
concurrent setting that must deal with non-trivial coordination and
synchronization costs, even without taking distribution or replication issues
into account.

\name shares resemblance to conflict-free replicated data types
(CRDT)~\cite{crdt}. CRDTs define abstract data types such as counters, sets,
etc., with commutative operations such that the state of the data type always
converges. Unlike CRDTs, the operations on mergeable types in \name need not
commute and the reconciliation protocol is defined by user-defined merge
functions. Moreover, CRDTs are not composable and each CRDT tends to be built
from scratch over the network protocol. Compare to this, \name types are
composable; mergeable ropes are polymorphic over its mergeable content. \name
uses 3-way merges using the lowest common ancestor, which is critical for all
of our user-defined merges. However, CRDTs do not have the benefit of lowest
common ancestor for merges and are only presented with the two concurrent
versions. If a 3-way merge is desired, then the causal history has to be
explicitly encoded in the data type. As a result, constructing even simple data
types like counters are more complicated using CRDTs~\cite{crdt} compared to
their implementation in \name.

>>>>

There have been numerous proposals over the years for realizing
distributed programming functionality in a functional programming
context.  Erlang~\cite{Armstrong}, which is the most well-known, is
based on a ``shared-nothing'' message-passing programming methodology.
Poly/ML~\cite{Mat97} and Facile~\cite{TLK+93} extend Standard ML,
while Acute~\cite{SLW+05} and HashCaml~\cite{BSS+06} extend OCaml,
with distributed programming abstractions and support, with particular
focus on type-safe marshalling and mobility.  Kali-Scheme~\cite{CJK95}
defines a distributed extension for Scheme-48, and Cloud
Haskell~\cite{EBPJ11} describes a domain-specific language that
enables development of distributed Haskell
programs. Obliq~\cite{Car95} explores issues related to mobility,
lexical scoping, and network references in an object-oriented
framework that enables safe (i.e., lexical) transmission of
closures. Orleans~\cite{Orleans} provides persistent actor-based
programming model for elastic cloud application, but the focus is on
strong consistency, mirroring a single-threaded execution.

However, unlike \name, none of these systems consider issues of object
replication or consistency as central to their programming models. The
realities of modern-day scalable cloud-based distributed systems,
especially those that are constructed from geographically distributed
participants, dictate that we must necessarily grapple with
consistency issues to achieve any kind of reasonable performance; it
is noteworthy that replication is inherent in many widely-used
cloud-based distributed system implementations today (e.g.,
Cassandra~\cite{Cassandra} or Dynamo~\cite{Dynamo}), which provide
only weak consistency guarantees.  Rather than exposing a low-level
operational treatment of replication, foisting the burden of reasoning
about replicated state onto the programmer, our primary contribution
is the development of a principled high-level approach to framing
these issues that abstracts these kinds of system-level details.  Our
goal is to leverage functional programming principles to minimize
disruption to the way programmers structure and reason about their
applications, without compromising efficiency or performance.  Indeed,
the semantics of mergeable types and versioning has utility in any
concurrent setting that must deal with non-trivial coordination and
synchronization costs, even without taking distribution or replication
issues into account.

Finally, \name shares some resemblance to conflict-free replicated
data types (CRDT)~\cite{crdt}. CRDTs define abstract data types such
as counters, sets, etc., with commutative operations such that the
state of the data type always converges. Unlike CRDTs, the operations
on mergeable types in \name need not commute and the reconciliation
protocol is defined by user-defined merge functions. Moreover, CRDTs
are not composable and each CRDT tends to be built from scratch over
the network protocol. Compared to this, \name types are composable;
mergeable ropes are polymorphic over its mergeable content. \name uses
3-way merges using the lowest common ancestor, which is critical for
all of our user-defined merges. However, CRDTs do not have the benefit
of lowest common ancestor for merges and are only presented with the
two concurrent versions. If a 3-way merge is desired, then the causal
history has to be explicitly encoded in the data type. As a result,
constructing even simple data types like counters are more complicated
using CRDTs~\cite{crdt} compared to their implementation in \name.
